Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-03 main.tex
Re-wrote abstract and part of intro.

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. We demonstrate
through extensive simulation and real-world experiment that the proposed
algorithms significantly speed up the sparsity recovery problem compared to the
state of the art.  Despite the low arithmetic intensity of ALM solvers, the GPU
can still outperform the CPU by a wide margin for large problem sizes, and when
solving many small problems simultaneously. The CPU implementation remains
faster for solving a single moderately sized problem. The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of users or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target Nvidia's CUDA GPU architecture
as well as the x86-64 architecture, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
CUDA programmable GPU architecture, which is also used in this paper.

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have big performance implications.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance implications to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. 
%GPUs are commonly shipped with a
%single GPU chip per PCI-express card.  Note that while typical engineering
%workstations can only support a single high-end GPU due to their high current
%draw, purpose-built servers can support multiple high-end GPU's.  
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.

@@

Our target GPU architecture is the Nvidia Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

On the GPU, most of the operations map well onto similar calls in Nvidia's GPU
BLAS library (CUBLAS) which, like MKL, can take advantage of both levels of
concurrency available in the hardware architecture.  Operations that do not map
well onto the CUBLAS API were implemented as GPU kernels (in C for CUDA).
Additionally, some operations that could have been implemented via multiple
BLAS calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.

>>>>

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
Due to the low arithmetic intensity of ALM solvers, both
architectures are bandwidth limited.  
with large problems running faster on the GPU, and small problems running faster
on the CPU.  
On both architectures, the proposed implementations dramatically out-perform
nieve library-based implementations. 
At the problem sizes most important for $\ell_1$-min based
face recognition, both architectures achieve comparable performance.
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of users or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target NVIDIA's CUDA GPU architecture
as well as the x86-64 architecture, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on 
the same GPU architecture addressed in this paper.

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

Our discussion and experiments will focus on the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred users with contemporary hardware
would require a server(or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.\footnote{Non-shared memory configurations
include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

Our target GPU architecture is the NVIDIA Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

On the GPU, most of the operations map well onto similar calls in NVIDIA's GPU
BLAS library (CUBLAS) which, like MKL, can take advantage of both levels of
concurrency available in the hardware architecture.  Operations that do not map
well onto the CUBLAS API were implemented as GPU kernels (in C for CUDA).
Additionally, some operations that could have been implemented via multiple
BLAS calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.
